Identifying Shifts in the Share Distribution of Transboundary Fish Stocks to Lead Time for Policy Response Under Climate Change

### 2021-02-16 Scripts/Manuscript.Rmd
Changed discussion and added supplemental fishbase table

---
output:
  bookdown::pdf_book:
    toc : no
    keep_tex: no
    number_sections: yes
geometry: margin=1in
editor_options:
  chunk_output_type: inline
bibliography: ../References/ToE_Ref.bib
header-includes:
- \usepackage{amsmath}
- \usepackage{float} # Use the 'float' package
- \floatplacement{figure}{H} # Make every figure with caption = h
---

@@

#### Project's Library
packages <- c(
  "tidyverse" ,
  "png",
  "grid",
  "knitr"#,
)

@@

@@

Over the last century, human activities have altered the physical and biogeochemical conditions of the ocean, including warming, acidification and reducing oxygen content [@IPCC:2019tn]. Distributions of marine species are closely related to the environment and the species’ preferences to environmental conditions (e.g., temperature, oxygen, salinity) [@Hutchinson:1957nt; @Nelson:2016kk]). As a result of climate change, many marine species have changed their distributions towards higher latitude, deeper water or following local temperature gradients [@Poloczanska:2016kk]. Biogeography of marine species is projected to continue to shift as ocean conditions are changing in the $21^{st}$ century under climate change [@Cheung:2010dt], impacting fisheries production and compromising our capacity to reach international sustainability goals such as Sustainable Development Goal 14 - life below water [@UnitedNations2018; @Barange:2014kr; @Singh:2017ds]. The projected risks of impacts can be reduced through improving the effectiveness of current fisheries management [@Gaines:2018sg; @Cheung:2018dh], including the fisheries management of species that cross international borders, i.e., shared stocks [@Gaines:2018sg; @Pinsky:2018cb].

The concept of shared stocks was developed following the ratification of the United Nations Convention on the Law of the Sea (UNCLOS) and the claiming of Exclusive Economic Zones (EEZs) by States [@UnitedNations:1986tl]. As defined by the United Nations’ Food and Agriculture Organization (FAO), shared stocks can be classified into four non-exclusive categories: (i) transboundary stocks, those that cross neighboring EEZs; (ii) straddling stocks, that, in addition to neighboring EEZs, also visit the adjacent high seas; (iii) highly migratory stocks, mainly tunas and bill-fishes, that migrate across vast oceanic regions including both the high seas and EEZs; and finally (iv) discrete stocks that are only present in the high seas [@Munro:2004th]. This study focuses on transboundary stocks exploited by fisheries operating within EEZs. While countries are responsible for the management of stocks within their EEZs, under UNCLOS, States are encouraged to cooperate when stocks are shared [@UnitedNations:1986tl]. A recent study estimates that, globally, there are 633 transboundary fish species representing 67% or the identified fished taxa, yielding an annual average of 48.5 million tonnes of catch and USD 78 billion in fishing revenue between 2005 and 2010 (Chapter 2).

The effectiveness of fisheries management for transboundary stocks is challenged by the shifts in marine stocks distribution under climate change [@Pinsky:2018cb; @Pinsky:2014by]. In many cases, catch or fishing effort quotas for transboundary stocks are based on historical records [@Baudron:2020dj] and do not necessarily consider the biogeography of the stocks [@FredstonHermann:2018kp], nor the effects of climate change on the fish stocks and fisheries [@PalaciosAbrantes:2020tv; @Sumby:2021gw]. Misalignment between fisheries resources allocation and stocks’ distributional shifts have previously resulted in unsustainable harvest and international disputes [@Miller:2013iv; @Song:2017va; @Spijkers:2017ij], patterns that are expected to be exacerbated by intensifying climate change [@Pinsky:2018cb]. With future shifts in species distributions under climate change, the challenges for the fisheries management of transboundary stocks will continue to increase during the 21st century. In order to prepare ocean governance to meet the resource management challenges arising from shifting transboundary stocks, it is important to constrain when climate change will affect the dynamics of transboundary  stocks and the intensity of the resulting impacts [@Pinsky:2018cb; @PalaciosAbrantes:2020tv; @Link:2010ei].

The timing and intensity of climate impacts on managing transboundary fish stocks can be quantitatively examined through the concepts of ‘time of emergence’ (ToE) and ‘threat point’ (TP). Time of emergence is defined as the moment in time when a signal (e.g., future anthropogenic trend) emerges from the background noise of natural variability (e.g., historical natural variation) [@Hawkins:2012hc] and has been applied to multiple oceanic physical and biogeochemical variables [@Rodgers:2015jq; @Keller:2014jv; @Frolicher:2016dg; @Cheung:2020ja; @Schlunegger:2020gz; @Schlunegger:2019el; @Henson:2017ev]. The premise behind the time of emergence is that we can only be confident that a significant change has been detected when the signal of anthropogenic climate change is larger than the background noise of natural climate variability [@Hawkins:2012hc]. The concept of threat point comes from game theory and is defined as the minimum payoff a player is willing to receive to cooperate in a game theoretic model [@Nash:1953vr]. Game theory has been widely used to investigate and manage transboundary stocks [@Sumaila:2013hv], including the impacts of climate change on the economics of transboundary stocks between Canada and the United States (Sumaila et al., *in press*) .

Here, I estimate the time of emergence and threat point of the share distribution of transboundary stocks worldwide. I employ a species distribution model driven by output from a comprehensive Earth System Model with ten ensemble members to project the distribution of 663 transboundary species (Chapter 2) across 280 EEZs of 198 coastal countries under a high greenhouse gas emissions scenario RCP 8.5 (see Methods). I treated each species in an EEZ as a single stock due to the lack of more spatially resolved data to delineate the boundary of a population (Chapter 2) and only considered shared stocks between neighboring EEZs (i.e., transboundary stocks that exclude  the high seas. I find that many transboundary stocks have already ‘emerged’ from historical natural variability and exceeded their share threat point in most neighboring EEZs, posing a threat to their sustainability and the resilience of international treaties.

@@

This analysis includes 633 exploited marine transboundary species previously identified to account for 80% of the catch taken from the world’s EEZs between 2005 and 2014 (Chapter 2). Each time a species was shared by a pair of neighbouring EEZs it was considered a transboundary stock [@Teh:2015gd, Chapter 2], resulting in a total of 9,132 transboundary stocks in this analysis. Using the transboundary species Atlantic cod (*Gadus morhua*) as an example, and based on the definition of transboundary stock in this study, the United States and Canada share a stock of Atlantic cod, and Canada and Greenland share another stock of Atlantic cod, but the United States and Greenland do not share a stock of this transboundary species. I defined the boundaries of the world’s EEZs using the Sea Around Us spatial division (updated 1 July 2015, available from http://www.seaaroundus.org), noting that it subdivides the EEZs of 198 coastal states into 280 regions (Figure \@ref(fig:toeFigureS1) including island territories. I determined the intersections between polygons using the R-statistical software package sf [@PackagesfSimpleF:2018vp]. Each EEZ was categorized by geopolitical region according to the United Nations (https://population.un.org/wpp/DefinitionOfRegions/) and biome [@Reygondeau:2019uh]. The habitat association of each species was determined following the classification in FishBase for fish species (http://www.fishbase.org) and SeaLifeBase for invertebrates (http://www.fishbase.org). For each stock and EEZ, I used the Sea Around Us data to estimate catch and fishing revenue from fishing activities within global EEZs. I report both average catch and revenue for the last decade (2005-2014) [@Zeller2016; @Tai2017; @Sumaila:2015uc]. The monthly average US Consumer Price Index (CPI) according to the U.S. Bureau of Labor Statistic (https://www.bls.gov/cpi/) was used to standardize all monetary values to 2019 real USD.

@@

We projected the distribution of marine species using a dynamic bioclimatic envelope model (hereafter called DBEM) [@Cheung:2010dt; @Cheung:2016jd]. The DBEM represents species’ physiology, habitat suitability, depth and latitudinal ranges, and spatial population dynamics as well as preferences for sea temperature, salinity, oxygen content, sea ice extent (for polar species) and bathymetry. For pelagic species, the model uses environmental variables at the surface whereas demersal species distribution is driven by ocean bottom variables . DBEM then estimates species abundance and maximum catch potential (a proxy of maximum sustainable yield) over a $0.5^{\circ}$ x $0.5^{\circ}$ grid (see Cheung et al. [-@Cheung:2010dt; -@Cheung:2016jd] for model details). Importantly, DBEM is able to project catches for the world’s large marine ecosystems and changes in catches by Exclusive Economic Zones that are consistent with observational-based estimates of catch from 1950 to 2016 [@Cheung:2016jd, Cheung et al. *in revision*].

The DBEM was forced with simulated ocean conditions from a ten-ensemble member simulation of the Geophysical Fluid Dynamics Laboratory Earth system model (GFDL-ESM2M) to project the distributions of the 633 species from 1951 to 2100 [@Dunne:2012bo; @Dunne:2013db; @Rodgers:2015jq]. The GFDL-ESM2M is run under historical forcing until 2005 and follows the high greenhouse gas emissions scenario, the Representative Concentration Pathway 8.5 (RCP 8.5) over the 2006-2100 period [@Riahi:2011dk]. Because the main approach of this paper relies on understanding the spatial and temporal variation of a stocks’ distribution, I have to understand distribution variability during both the historical and the future periods, to infer differences between time frames. I do this by following an ensemble member approach where each member constitutes a different realization of the Earth system variability condition [@Rodgers:2015jq; @Frolicher:2009ge]. Thus, for my experiment, each of the ten GFDL-ESM2M ensembles were started from infinitesimally small differences in Earth system initial conditions in year 1950 resulting in a unique atmosphere and ocean state at each point in time after about three years for surface waters and eight years for subsurface waters [@Frolicher:2020cc]. By design, variations among ensemble members are then solely due to natural variability. This approach allows us to estimate the natural internal variability (e.g., background noise) and isolate the forced anthropogenic climate signal of a stock’s distribution by averaging the secular trend over all ten ensemble members.

@@

I developed an index to evaluate range shifts in transboundary stocks under climate change (TI). This index is based on the shifts in the distribution centroid of a transboundary stock relative to the centroid of the neighboring EEZs that share this stock (Figure \@ref(fig:toeFigureS2). The centroid of a transboundary stock was determined by the average (µ) latitude ($lat_{ts}$) and longitude ($lon_{ts}$) across the grid cells with the highest abundance within the neighboring EEZs sharing the stock e.g., ^Pacific halibut (*Hippoglossus stenolepis*) that is transboundary between the United States and Canada. Therefore,

@@

where $lat_{per}$ and $lon_{per}$ are the latitudes and longitudes of the grid cells holding the $per^{th}$ percentile of the projected transboundary stock abundance. To focus on areas where transboundary stocks are more abundant and fishing activities are more likely to take place, I included grid cells where the projected stock abundance within the neighbouring EEZs sharing the stock was above the top 95th percentile (e.g., per = 95%). A sensitivity analysis was ran using a subset of species (n = 34) for all EEZs (per = 20th, 50th and $90^{th}$ percentile) and examined the effects of different thresholds on the calculated index value (Figure \@ref(fig:toeFigureS3)). The centroid of each EEZ was estimated using the st package in R (Figure \@ref(fig:toeFigureS1)). For each ensemble member, neighboring EEZs and transboundary stock, I computed the distance between centroids assuming the earth is a perfect sphere and ignoring geographic barriers using the geosphere package in R;

@@

where $D_A$ and $D_B$ represent the distance between the distribution centroids of a stock and the centroid of the pair of neighbouring EEZs (A and B) sharing the stock for each time step from 2006 to 2100 (*t*); and *sd* is the standard deviation of the historical (*th*, 1951 - 2005) centroid distribution for $D_A$ and $D_B$. Thus, a higher TI means that a shared stock becomes more or less abundant relative to its neighbouring EEZs that is beyond what the fisheries experienced historically. From the perspective of a state’s fisheries, higher TI indicates an increase in the challenges to effective management and sustainability of the shared stock’ fisheries. In contrast, a smaller TI suggests that the sharing of the stocks between neighbouring EEZs is becoming relatively more stable; a condition that favors effective management and sustainability of the shared stocks’ fisheries. The index was smoothed to a 10-year average to reduce interannual variability.

@@

\begin{equation}
\begin{split}
\label{eq:four}
EmergenceTIndex = \frac{\mu_{TI[t]}}{SD_{TI_{th}}} \\
ToE = ToE_{t} > 1 s.d.
\end{split}
\end{equation}

where $\mu_{TI}$ is the yearly mean TI ($\mu$) across all ten ensemble members in year *yr* and $SD_{TI}$ is the standard deviation of the smoothed (10-year moving average) TI across the ensemble members for the historical reference period (1951 - 2005). I set a threshold of TI above which I consider a shared stock to have emerged from the historical internal variability of the TI. The threshold was set as one standard deviation of the variability ($SD_{TI}$) (i.e., Equation \@ref(eq:four), representing a probability of 68% that the transboundary index has emerged from historical variability). I tested the sensitivity of the calculated emergence index to  a higher emergence threshold of two times the standard deviation (i.e., representing a probability of 95% that the index has emerged).

@@

<!-- I created a country-specific vulnerability index (VI) to examine the risk of impacts of species’ range shifts on transboundary fisheries management. The VI reflects both the rate of change of shared stocks’ distribution (indicated by the time-of-emergence of the transboundary index) and the importance of the shifting species to a country’s total catch and revenue from the species. The VI is calculated from: -->

@@

In game theory, cooperation over a shared resource will more likely result in a better overall outcome for the sharing parties than those operating on a non-cooperative basis [@Munro:1979dg]. However, the benefits that each player gets from a cooperative strategy must be above a minimum payoff i.e., “threat point” [@Sumaila:2013hv]. Here, I define the threat point as the minimum required proportion of a shared stock present within an EEZ for a country to engage in cooperative management with their sharing neighbor [@PalaciosAbrantes:2020vt; @Sumaila:2020tp]. Any proportion below the defined threat point of a country would result in a unilateral management of the shared stock.

First, I estimated the stock share ratio (SSR) of each transboundary stock shared by neighbouring nations between 1951 and 2100. I did this by aggregating the number of $0.5^{\circ}$ x $0.5^{\circ}$ grid cells in which the stock is present across neighboring EEZs, and then calculated the proportion of the stock held within each EEZ (Chapter 2). Second, I averaged the calculated proportion of stock within an EEZ held into three time periods to reduce the effects of variability. The first time period ($SSR_{th}$) spans 1951 to 2005 and is chosen to match the historical period in the GFDL ESM2M simulations, over which the model was forced with observation-based greenhouse gas, aerosol and natural external forcing [@Dunne:2012bo; @Dunne:2013db]. It is assumed to represent the historic baseline of shared distributions for each stock. I then selected two future periods; the early $21^{st}$ century as the average of 2021 to 2040 ($SR{te}$), and the mid $21^{st}$ century as the average of 2041 to 2060 ($SSR{tm}$). I chose these time periods to match the challenges of achieving fisheries-related UN-SDGs (6) such as SDG 14.4 (end overfishing), SDG 2.4 (ensure sustainable food production systems) or SDG 1.2 (poverty reduction), to be achieved by 2030 [@Singh:2017ds]. The analysis was replicated for projected stocks distributions from each of the ten ensemble members and results were averaged across ensemble members. Third, I defined a threat point for each EEZ’s stock as $SSR_{th} \pm \sigma$, where $\sigma$ is the standard deviation of $SSR_{th}$. This way, a change of SSR beyond an EEZ’s threat point happens when the future SSR exceeds one standard deviation of the historical variations of the SSR i.e when $SSR_{et} \geqq (SSR_{th} + \sigma)$ or $SSR_{e,m} \leqq (SSR_{th} - \sigma)$. Finally, I estimated the percentage change in the SSR ($\Delta{SSR}$) of each future time period ($SSR_f$) relative to the historic time period ($SSR_{th}$) for each stock whose share ratio overpassed the threat point following Palacios-Abrantes et al. (*in press*),

@@

My analysis suggests that most (n = 47) countries responsible for 75% of the yearly revenue generated from transboundary stocks between 2005 and 2010 will see the distribution range of their stocks shift beyond historical variabilities between 2025 and 2040 (Figure \@ref(fig:toeFigure3)). In some cases, like Indonesia (ranked $8^{th}$ in terms of revenue from transboundary stocks), Thailand (ranked $7^{th}$ in terms of revenue from transboundary stocks) and the United States (ranked $4^{th}$ in terms of fishing revenue from transboundary stocks), the range shift of transboundary stocks will emerge by around 2030. The 47 countries that make up the bulk of the global fisheries revenues also have more transboundary stocks with range shifts emerging from historical variability, such as Spain (n = 122), France (n = 85), China and Portugal (both n = 73) and Senegal (n = 66), which are the top 5 leading countries in terms of number of transboundary stocks emerging from historical variability.

>>>>

---
title: Identifying Shifts in the Share Distribution of Transboundary Fish Stocks to Lead Time for Policy Response Under Climate Change
subtitle: Challenges to transboundary fisheries under climate change
author: 'Authors: Juliano Palacios-Abrantes*^1^, Thomas Frölicher^2,3^, Gabriel Reygondeau^1^, U. Rashid Sumaila^1^, Alessandro Tagliabue^4^, Colette Wabnitz^1,5^, and William W.L. Cheung^1^'
date: 'Affiliations: ^1^Institute for the Oceans and Fisheries, University of British Columbia, Vancouver, Canada. ^2^Climate and Environmental Physics, Physics Institute, University of Bern. ^3^Oeschger Centre for Climate Change Research, University of Bern, Switzerland. ^4^School of Environmental Sciences, University of Liverpool. ^5^Center for Ocean Solutions, Stanford University, Stanford, United States'
output:
  # word_document:
    # reference_docx: ../References/template.docx
  pdf_document: default
bibliography: ../References/ToE_Ref.bib
csl: ../References/science.csl
header-includes:
- \usepackage{amsmath}
- \usepackage{float} # Use the 'float' package
- \floatplacement{figure}{H} # Make every figure with caption = h
---

<!-- # Instructions  -->

<!-- Research articles should present a major advance and can be short or long (maximum 15,000 words) format. Submissions must include an abstract (up to 150 words), an introduction, and sections with brief informative subheadings. You may include up to ten figures and/or tables and about 60 references. Supplementary materials should be limited to information that is not essential for the general understanding of the research presented in the main text and can include data sets, figures, tables, videos, or audio files.  You may include up to a total of ten figures and/or tables (combined) throughout the supplemental text.  -->

@@

#### Project's Library
packages <- c(
  "tidyverse" ,
  "png",
  "grid",
  "knitr",
 "kableExtra"
  # "here", # for dbem_import `here()`
  # "data.table", #dbem_import `fread()`
  # "readxl", # for reading excell files
  # "janitor", # for clearing names
  # "geosphere", # estimate distances between points `distm()`
  # "ggrepel",
  # "zoo", # for average mean
  # "parallel", # for mclapply,
  # "sf", # for mapping
  # "st", # for mapping
  # "rgdal", #Spatial analysis
  # "tools", #Spatial analysis 
  # "zeallot", # for Juanito's map
  # "gmt", # for estimating distances between points
  # "viridis"
)

@@

<!-- * corresponding author: j.palacios@oceans.ubc.ca -->

<!-- # Abstract (150  words) -->

<!-- An opening sentence that states the question/problem addressed by the research AND Enough background content to give context to the study AND A brief statement of primary results AND A short concluding sentence. -->

Climate change shifts in marine species distributions is affecting the current frameworks emplaced to manage species shared fisheries. Here, we rely on a modeling framework with multiple ensemble members to identify the time frame by which shared stocks distribution will change from historical ranges, and the change in the proportion of the distrirbution that each EEZ has under climate change. Our results suggest that 60% of shared species have emerged since 2006 in 87% of the world EEZs. By 2075 all EEZs will have an emerging species and by 2097 and all species will have emerged in at least one EEZ. Average changes in the ratio of species-share between neighboring EEZs will fluctuate between 1 and 10% with some exceeding 25% by 2030 relative to today’s ratio. Our results suggest that most changes will happen before 2030 urging policy to be re-formulated to account for the possible implications of these changes.

@@

Over the last century, human activities have altered the physical and biogeochemical conditions of the ocean, including warming, acidification and reducing oxygen content [@IPCC:2019tn]. Distributions of marine species are closely related to the environment and the species’ preferences to environmental conditions (e.g., temperature, oxygen, salinity) [@Hutchinson:1957nt; @Nelson:2016kk]. As a result of climate change, many marine species have changed their distributions towards higher latitude, deeper water or following local temperature gradients [@Poloczanska:2016kk]. Biogeography of marine species is projected to continue to shift as ocean conditions are changing in the $21^{st}$ century under climate change [@Cheung:2010dt], impacting fisheries production and compromising our capacity to reach international sustainability goals such as Sustainable Development Goal 14 - life below water [@UnitedNations2018; @Barange:2014kr; @Singh:2017ds]. The projected risks of impacts can be reduced through improving the effectiveness of current fisheries managemen [@Gaines:2018sg; @Cheung:2018dh], including the fisheries management of species that cross international borders, i.e., shared stocks [@Gaines:2018sg; @Pinsky:2018cb].

The concept of shared stocks was developed following the ratification of the United Nations Convention on the Law of the Sea (UNCLOS) and the claiming of Exclusive Economic Zones (EEZs) by States [@UnitedNations:1986tl]. As defined by the United Nations’ Food and Agriculture Organization (FAO), shared stocks can be classified into four non-exclusive categories: (*i*) transboundary stocks, those that cross neighboring EEZs; (*ii*) straddling stocks, that, in addition to neighboring EEZs, also visit the adjacent high seas; (*iii*) highly migratory stocks, mainly tunas and bill-fishes, that migrate across vast oceanic regions including both the high seas and EEZs; and finally (*iV*) discrete stocks that are only present in the high seas [@Munro:2004th]. This study focuses on transboundary stocks exploited by fisheries operating within EEZs. While countries are responsible for the management of stocks within their EEZs, under UNCLOS, States are encouraged to cooperate when stocks are shared [@UnitedNations:1986tl]. A recent study estimates that, globally, there are 633 transboundary fish species representing 67% or the identified fished taxa, and yielding an annual average of 48.5 million tonnes of catch and USD 78 billion in fishing revenue between 2005 and 2010 [@PalaciosAbrantes:2020tv].

The effectiveness of fisheries management for transboundary stocks is challenged by the shifts in marine stocks distribution under climate change [@Pinsky:2018cb; @Pinsky:2014by]. In many cases, catch or fishing effort quotas for transboundary stocks are based on historical records [@Baudron:2020dj] and do not necessarily consider the biogeography of the species [@FredstonHermann:2018kp], nor the effects of climate change on the fish stocks and fisheries [@PalaciosAbrantes:2020tv; @Sumby:2021gw]. Misalignment between fisheries resources allocation and species’ distributional shifts have previously resulted in unsustainable harvest and international disputes [@Miller:2013iv; @Song:2017va; @Spijkers:2017ij], patterns that are expected to be exacerbated by intensifying climate change [@Pinsky:2018cb]. With future shifts in species distributions under climate change, the challenges for the fisheries management of transboundary stocks will continue to increase during the $21^{st}$ century. In order to prepare ocean governance to meet the resource management challenges arising from shifting transboundary stocks, it is important to constrain when climate change will affect the dynamics of transboundary stocks and the intensity of the resulting impacts [@Pinsky:2018cb; @PalaciosAbrantes:2020tv; @Link:2010ei].

The timing and intensity of climate impacts on managing transboundary fish stocks can be quantitatively examined through the concepts of ‘time of emergence’ (ToE) and ‘threat point’ (TP). Time of emergence is defined as the moment in time when a signal (e.g., future anthropogenic trend) emerges from the background noise of natural variability (e.g., historical natural variation) [@Hawkins:2012hc] and has been applied to multiple oceanic physical and biogeochemical variables [@Rodgers:2015jq; @Keller:2014jv; @Frolicher:2016dg; @Cheung:2020ja; @Schlunegger:2020gz; @Schlunegger:2019el; @Henson:2017ev]. The premise behind the time of emergence is that we can only be confident that a significant change has been detected when the signal of anthropogenic climate change is larger than the background noise of natural climate variability [@Hawkins:2012hc]. The concept of threat point comes from game theory and is defined as the minimum payoff a player is willing to receive to cooperate in a game theoretic model [@Nash:1953vr]. Game theory has been widely used to investigate and manage transboundary stocks [@Sumaila:2013hv], including the impacts of climate change on the economics of transboundary species between Canada and the United States [@Sumaila:2020tp].

Here, we estimate the time of emergence and threat point of the share distribution of transboundary stocks worldwide. We employ a species distribution model driven by output from a comprehensive Earth System Model with ten ensemble members to project the distribution of 663 transboundary species [@PalaciosAbrantes:2020tv] across 280 EEZs of 198 coastal countries under a high greenhouse gas emissions scenario RCP8.5 (see Methods). We treated each species in an EEZ as a single stock due to the lack of more spatially resolved data to delineate the boundary of a population [@PalaciosAbrantes:2020tv] and only considered shared stocks between neighboring EEZs (i.e., transboundary stocks that exclude the high seas). We find that many transboundary stocks have already ‘emerged’ from historical natural variability and exceeded their share threat point in most neighboring EEZs, posing a threat to their sustainability and the resilience of international treaties.

@@

This analysis includes 633 exploited marine transboundary species previously identified to account for 80% of the catch taken from the world’s EEZs between 2005 and 2014 [@PalaciosAbrantes:2020tv]. Each time a species was shared by a pair of neighbouring EEZs it was considered a transboundary stock [@PalaciosAbrantes:2020tv]; @Teh:2015gd], resulting in a total of 9,132 transboundary stocks in this analysis. Using the transboundary species Atlantic cod (*Gadus morhua*) as an example, and based on the definition of transboundary stock in this study, the United States and Canada share a stock of Atlantic cod, and Canada and Greenland share another stock of Atlantic cod, but the United States and Greenland do not share a stock of this transboundary species. We defined the boundaries of the world’s EEZs using the Sea Around Us spatial division (updated 1 July 2015, available from http://www.seaaroundus.org), noting that it subdivides the EEZs of 198 coastal states into 280 regions (Figure \@ref(fig:toeFigureS1)) including island territories. We determined the intersections between polygons using the R-statistical software package sf [@PackagesfSimpleF:2018vp]. Each EEZ was categorized by geopolitical region according to the United Nations (https://population.un.org/wpp/DefinitionOfRegions/) and biome [@Reygondeau:2019uh]. The habitat association of each species was determined following the classification in FishBase for fish species (http://www.fishbase.org) and SeaLifeBase for invertebrates (http://www.fishbase.org). For each stock and EEZ, we used the Sea Around Us data to estimate catch and fishing revenue from fishing activities within global EEZs. We report both average catch and revenue for the last decade (2005-2014) [@Zeller2016; @Tai2017; @Sumaila:2015uc]. The monthly average US Consumer Price Index (CPI) according to the U.S. Bureau of Labor Statistic (https://www.bls.gov/cpi/) was used to standardize all monetary values to 2019 real USD.

@@

We projected the distribution of marine species using a dynamic bioclimatic envelope model (hereafter called DBEM) [@Cheung:2010dt; @Cheung:2016jd]. The DBEM represents species’ physiology, habitat suitability, depth and latitudinal ranges, and spatial population dynamics as well as preferences for sea temperature, salinity, oxygen content, sea ice extent (for polar species) and bathymetry. For pelagic species, the model uses environmental variables at the surface whereas demersal species distribution is driven by ocean bottom variables . DBEM then estimates species abundance and maximum catch potential (a proxy of maximum sustainable yield) over a $0.5^{\circ}$ x $0.5^{\circ}$ grid (see Cheung et al. 2010 and 2016 for model details). Importantly, DBEM is able to project catches for the world’s large marine ecosystems and changes in catches by Exclusive Economic Zones that are consistent with observational-based estimates of catch from 1950 to 2016 [@Cheung:2016jd], Cheung et al. *in revision*].

We forced the DBEM with simulated ocean conditions from a ten-ensemble member simulation of the Geophysical Fluid Dynamics Laboratory Earth system model (GFDL-ESM2M) to project the distributions of the 633 species from 1951 to 2100 [@Dunne:2012bo; @Dunne:2013db; @Rodgers:2015jq]. The GFDL-ESM2M is run under historical forcing until 2005 and follows the high greenhouse gas emissions scenario, the Representative Concentration Pathway 8.5 (RCP 8.5) over the 2006-2100 period [@Riahi:2011dk]. Because the main approach of this paper relies on understanding the spatial and temporal variation of a stocks’ distribution, we have to understand distribution variability during both the historical and the future periods, to infer differences between time frames. We do this by following an ensemble member approach where each member constitutes a different realization of the Earth system variability condition [@Rodgers:2015jq; @Frolicher:2009ge]. Thus, for our experiment, each of the ten GFDL-ESM2M ensembles were started from infinitesimally small differences in Earth system initial conditions in year 1950 resulting in a unique atmosphere and ocean state at each point in time after about three years for surface waters and eight years for subsurface waters [@Frolicher:2020cc]. By design, variations among ensemble members are then solely due to natural variability. This approach allows us to estimate the natural internal variability (e.g., background noise) and isolate the forced anthropogenic climate signal of a stock’s distribution by averaging the secular trend over all ten ensemble members.

@@

I developed an index to evaluate range shifts in transboundary stocks under climate change (TI). This index is based on the shifts in the distribution centroid of a transboundary stock relative to the centroid of the neighboring EEZs that share this stock (Figure \@ref(fig:toeFigureS2)). The centroid of a transboundary stock was determined by the average (µ) latitude ($lat_{ts}$) and longitude ($lon_{ts}$) across the grid cells with the highest abundance within the neighboring EEZs sharing the stock e.g., sockeye salmon (*Oncorhynchus nerka*) that is transboundary between the United States and Canada. Therefore,

@@

where $lat_{per}$ and $lon_{per}$ are the latitudes and longitudes of the grid cells holding the $per^{th}$ percentile of the projected transboundary stock abundance. To focus on areas where transboundary stocks are more abundant and fishing activities are more likely to take place, we included grid cells where the projected stock abundance within the neighbouring EEZs sharing the stock was above the  95th percentile (e.g., per = 95%). We ran a sensitivity analysis using a subset of species (n = 34) for all EEZs (per = 20th, 50th and $90^{th}$ percentile) and examined the effects of different thresholds on the calculated index value (Figure \@ref(fig:toeFigureS3)). The centroid of each EEZ was estimated using the st package in R (Figure \@ref(fig:toeFigureS1)). For each ensemble member, neighboring EEZs and transboundary stock, the distance between centroids was computed assuming the Earth is a perfect sphere and ignoring geographic barriers using the geosphere package in R;

@@

where $D_A$ and $D_B$ represent the distance between the distribution centroids of a species and the centroid of the pair of neighbouring EEZs (A and B) sharing the species for each time step from 2006 to 2100 (*t*); and *sd* is the standard deviation of the historical (*th*, 1951 - 2005) centroid distribution for $D_A$ and $D_B$. Thus, a higher TI means that a shared stock becomes more or less abundant relative to its neighbouring EEZs that is beyond what the fisheries experienced historically. From the perspective of a state’s fisheries, higher TI indicates an increase in the challenges to effective management and sustainability of the shared stock’ fisheries. In contrast, a smaller TI suggests that the sharing of the stocks between neighbouring EEZs is becoming relatively more stable; a condition that favors effective management and sustainability of the shared stocks’ fisheries. The index was smoothed to a 10-year average to reduce interannual variability.

@@

\begin{equation}
\begin{split}
\label{eq:four}
ETI = \frac{\mu_{TI[t]}}{SD_{TI_{th}}} \\
ToE = t when Emergence index[t] > 1 s.d.
\end{split}
\end{equation}

where $\mu_{TI}$ is the yearly mean TI across all ten ensemble members in year *yr* and $SD_{TI}$ is the standard deviation of the smoothed (10-year moving average) TI across the ensemble members for the historical reference period (1951 - 2005). We set a threshold of TI above which it was consider a shared stock to have emerged from the historical internal variability of the TI. The threshold was set as one standard deviation of the variability ($SD_{TI}$) (i.e., Equation \@ref(eq:four), representing a probability of 68% that the transboundary index has emerged from historical variability). We tested the sensitivity of the calculated emergence index to  a higher emergence threshold of two times the standard deviation (i.e., representing a probability of 95% that the index has emerged).

@@

<!-- We created a country-specific vulnerability index (VI) to examine the risk of impacts of species’ range shifts on transboundary fisheries management. The VI reflects both the rate of change of shared stocks’ distribution (indicated by the time-of-emergence of the transboundary index) and the importance of the shifting species to a country’s total catch and revenue from the species. The VI is calculated from: -->

@@

In game theory, cooperation over a transboundary resource will more likely result in a better overall outcome for the sharing parties than those operating on a non-cooperative basis [@Munro:1979dg]. However, the benefits that each player gets from a cooperative strategy must be above a minimum payoff i.e., “threat point” [@Sumaila:2013hv]. Here, we define the threat point as the minimum required proportion of a transboundary stock present within an EEZ for a country to engage in cooperative management with their sharing neighbor [@PalaciosAbrantes:2020vt; @Sumaila:2020tp]. Any proportion below the defined threat point of a country would result in a deflection from the joint management of the transboundary stock.

First we estimated the stock share ratio (SSR) of each transboundary stock shared by neighbouring nations between 1951 and 2100. We did this by aggregating the number of $0.5^{\circ}$ x $0.5^{\circ}$ grid cells in which the stock is present across neighboring EEZs, and then calculated the proportion of the stock held within each EEZ [@PalaciosAbrantes:2020tv]. Second, we average the calculated proportion of stock within an EEZ held into three time periods to reduce the effects of inter-annual climate variability. The first time period ($SSR_{th}$) spans from 1951 to 2005 and matches the historical period in the GFDLESM-2M simulations, over which the model was forced with observation-based greenhouse gas, aerosol and natural external forcing [@Dunne:2012bo; @Dunne:2013db]. It is assumed as the historic baseline of shared distributions for each stock. We then selected two future periods; the early $21^{st}$ century as the average of 2021 to 2040 ($SR{te}$), and the mid $21^{st}$ century as the average of 2041 to 2060 ($SSR{tm}$). We chose these time periods to match the challenges of achieving fisheries-related UN-SDGs (6) such as SDG 14.4 (end overfishing), SDG 2.4 (ensure sustainable food production systems) or SDG 1.2 (poverty reduction), to be achieved by 2030 [@Singh:2017ds]. The analysis was replicated for projected stocks distributions from each of the ten ensemble members and results were averaged across ensemble members. Third, we defined a threat point for each EEZ’s stock as $SSR_{th} \pm \sigma$, where $\sigma$ is the standard deviation of $SSR_{th}$. This way, a change of SSR beyond an EEZ’s threat point happens when the future SSR exceeds one standard deviation of the historical variations of the SSR i.e when $SSR_{et} \geqq (SSR_{th} + \sigma)$ or $SSR_{e,m} \leqq (SSR_{th} - \sigma)$. Finally, we estimated the percentage change in the SSR ($\Delta{SSR}$) of each future time period ($SSR_f$) relative to the historic time period ($SSR_{th}$) for each stock whose share ratio overpassed the threat point following Palacios-Abrantes et al. (*in press*),

@@

Our analysis suggests that most (n = 47) countries responsible for 75% of the yearly revenue generated from transboundary stocks between 2005 and 2010 will see their distribution range shifts beyond historical variabilities between 2025 and 2040 (Figure \@ref(fig:toeFigure3)). In some cases, like Indonesia (ranked $8^{th}$ in terms of revenue from transboundary stocks), Thailand (ranked $7^{th}$ in terms of revenue from transboundary stocks) and the United States (ranked $4^{th}$ in terms of fishing revenue from transboundary stocks), stocks’ range shifts swill emerge by around 2030. The 47 countries that make up the bulk of the global fisheries revenues also have more stocks with range shifts emerging from historical variabilities, being Spain (n = 122), France (n = 85), China and Portugal (both n = 73) and Senegal (n = 66) the top 5 leading countries in number of emerging stocks.
